---
title: "Notebook"
output: html_document
date: "2024-11-09"
---

**SET UP**

```{r}
rm(list = ls())

# Importing libraries
library("dplyr")
library("skimr")
library("ggplot2")
library("ggrepel")
library("stringi")
# Install.packages("fuzzyjoin")
library("fuzzyjoin")

```

We have different datasets which contain different statistics from different players which we will need to combine to obtain a richer ando more useful dataset.

```{r}
# Reading the different datasets
defense_stats <- read.csv("Big5-defenseStats22-23.csv", sep = ",")
gca_stats <- read.csv("Big5-gcaStats22-23.csv", sep = ",")
misc_stats <- read.csv("Big5-miscStats22-23.csv", sep = ",")
passing_stats <- read.csv("Big5-passingStats22-23.csv", sep = ",")
possession_stats <- read.csv("Big5-possessionStats22-23.csv", sep = ",")
shooting_stats <- read.csv("Big5-shootingStats22-23.csv", sep = ",")

#Passing types information is included in passing stats
# Csv statsStats has already included statistics

# Datasets for goalkeepers
keepers_one <- read.csv("Big5-keepersStats22-23.csv", sep = ",")

# Dataset for all players combined
playingtime <- read.csv("Big5-playingtimeStats22-23.csv", sep = ",")

```

**DATA PREPROCESSING**

Merging all datasets in one when each of them has a lot of datasets will potentially cause many problems in the long run, we won´t handle duplicates or NA values just yet and fill focus on reducing the amount of dimensions each dataset has.
Therefore, we will look at each dataset independently handling the unimportant columns or columns that can be inferred from other columns to reduce dimensionality.


```{r}
# Defense Stats

# The total amount of tackles is the sum of Def.3rd, Mid.3rd and Att.3rd which just refer to where those tackles were made, Tkl.Int is a sum of both tackles and interceptions therefore we can keep this columns and eliminate both interceptions and tackles. Blocks are a sum of Sh and Pass which refers to the different type of blocks. Tkl.1 is tackles in the first half, already included in the first variable. Tkl.1 measures the percentage of lost balls when Att is done, it divides Att and Lost

defense_cols <- c("Def.3rd","Mid.3rd","Att.3rd","Tkl","Tkl.1","Int","Sh", "Pass", "Att", "Lost")

defense_stats <- defense_stats %>% select(-all_of(defense_cols))


# GCA stats

# SCA is an advanced metric that tracks the two offensive/attacking actions that directly lead to a shot on # goal, Following the same logic as a Shot-Creating Action, a Goal-Creating Action is an advanced metric that tracks the two offensive/attacking actions that leads to a goal. These follow the same criteria as  the above list. The difference is that one causes a goald, the other an scoring chance

# SCA is a sum of PassLive, PassDead, TO, Sh Fld, Def, they are different stats that contribute to it, and  SCA90s can be inferred as SCA*90s returns SCA90s
# GCA follows the same deduction as before

gca_cols <- c("SCA90","PassLive", "PassDead", "TO", "Sh", "Fld", "Def", "GCA90", "PassLive.1", "PassDead.1", "TO.1", "Sh.1", "Fld.1", "Def.1")

gca_stats <- gca_stats %>% select(-all_of(gca_cols))



# Misc stats

# X2CrdY is when the player is suspended by 2 yellow cards instead of one red card, this is included in the Red Card statistic. TklW and Int are already included in the defense dataset, Won. refers to the aerial duels won, and it divides won and lost, so we can remove them. Own goals are many times fortunate and dont affect the skill value of a player. PKwon and PKcon refer to the amount of penalties conceded and won, but not many players have a value here so we will remove it as it is irrelevant. Crs refers to crosses and will appear in a future dataset 

misc_cols <- c("X2CrdY", "TklW", "Int", "Won", "Lost", "OG", "PKwon", "PKcon", "Crs")


misc_stats <- misc_stats %>% select(-all_of(misc_cols))

# Passing Stats

#Cmp. is the percentage of the completed passes in terms of the attempted ones, therefore we can remove them. Cmp.1, Att.1, Cmp..1,Cmp.2,Att.2, Cmp..2, Cmp.3,Att.3,Cmp..3 are all relative to the distance of the passes, which is not of much interest as of right now and both TotDist and PrgDist contain a lot of information about the distance of passes. 1/3, PPA, CrsPa, PrgP are also covered as they all refer to progressive passes

passing_cols <- c("Cmp.1", "Att.1", "Cmp..1","Cmp.2","Att.2", "Cmp..2", "Cmp.3","Att.3","Cmp..3","X1.3", "PPA", "CrsPA", "PrgP")

passing_stats <- passing_stats %>% select(-all_of(passing_cols))

# Posession Stats


# Touches is a sum of Def.Pen, Def.3rd, Mid.3rd, Att.3rd, Att.Pen, Live so we can remove them. Succ. is the division of Att and Succ which refers to the number of times the attacker has dribbled the defender divided by the total attemps. Tkld, Tkld. are two variables directly related to the previous so we can also take them out. X1.3, CPA, Mis are not really valuable. TotDist and PrgDist will have different names to differentiate from the previous dataset that used the same variables. Rec, PrgR refer to the times the player received a good pass where in most of the cases if a pass is missed it is usually the passers fault rather than the receiver.

possession_stats <- possession_stats %>% 
  rename(DistanceRan = TotDist, ForwardDistanceRan = PrgDist)

posession_cols <- c("Def.Pen", "Def.3rd", "Mid.3rd", "Att.3rd", "Att.Pen", "Live", "Att", "Succ", "Tkld", "Tkld.","X1.3", "CPA", "Mis", "Rec", "PrgR", "PrgC")

possession_stats <- possession_stats %>% select(-all_of(posession_cols))

# Shooting Stats

# With goals, shots and shots on target we can infer SoT., Sh.90, SoT.90, G.Sh, G.SoT. As not many players shoot penalties or free kicks we will not take into account this, FK, PK and PKatt will be removed

shooting_cols <- c("SoT.", "Sh.90", "SoT.90", "G.Sh", "G.SoT","FK", "PK","PKatt")

shooting_stats <- shooting_stats %>% select(-all_of(shooting_cols))

# Keepers one

# The statistics MP, Starts, Min will be in another dataset so we will remove them. Save. is calculated by SoTA-GA/SoTa and CS. which refers to clean sheets can also be removed. Everything about free kicks and penalties can be removed as it is easy to score from a penalty and harder from a free kick and it usually depends on the quality of the attacker, many times the keeper cannot do much.We will remove PKatt, PKA, PKsv,PKm,Save..1

keepers_cols <- c("MP", "Starts", "Min", "Save.", "CS.", "PKatt", "PKA", "PKsv","PKm","Save..1")

keepers_one <- keepers_one %>% select(-all_of(keepers_cols))

# Playing Time

# Mn.Mp can be calculated as Min/Mp, Mn.Start will be Min/Starts, Compl can be seen as it is not really important. The same for all sub variables as with minutes we can already kind of see the amount of time played. We will remove Subs, Mn.Sub, unSub

playing_cols <- c("Mn.MP", "Mn.Start", "Compl", "Subs", "Mn.Sub", "unSub")

playingtime <- playingtime %>% select(-all_of(playing_cols))

```




Merging all datasets according to different columns and dealing with missing values and duplicates later, this can be done with the inner join function taken into account that all datasets have the same 8 first columns.

```{r}
# Making a list with the repeated columns
repeated_columns = c("Player", "Nation", "Pos", "Squad", "Comp", "Age", "Born", "X90s")


# Inner joining all datasets based on key columns and a brief look at it
merged_players <- inner_join(defense_stats, gca_stats, by = repeated_columns)%>%
                  inner_join(misc_stats, by = repeated_columns)%>%
                  inner_join(passing_stats, by = repeated_columns)%>%
                  inner_join(possession_stats, by = repeated_columns)%>%
                  inner_join(shooting_stats, by = repeated_columns)

skim(merged_players)


```


We now have the problem that goalkeepers have different stats to players, therefore the number of variables varies and we cannot join them to the same dataset. Instead we need to add the columns which are not in players to goalkeepers with a 0 value for each entry and viceversa.

```{r}
# Obtaining all columns in players and goalkeepers
player_columns <- colnames(merged_players)
gk_columns <- colnames(keepers_one)

# Finding out the different columns in each of the datasets
missing_columns_players = setdiff(gk_columns, player_columns)
missing_columns_gk = setdiff(player_columns, gk_columns)

# Using a for loop to create the missing columns with a value of 0
for (col in missing_columns_players) {
  merged_players[[col]] <- 0
}

for (col in missing_columns_gk) {
  keepers_one[[col]] <- 0
}

# Combining both goalkeepers and players
all_players = bind_rows(merged_players,keepers_one)

# Merging with playing time stats
final_players <- inner_join(all_players, playingtime, by = repeated_columns)
head(final_players)

```

We observe many redundancies in this new dataset, such as that the nations are repeated in both lower and upper case, that some players have more than one position or that the competition also has lower case in front which could be troubling in the future. We will remove the strings with gsub and substr, and maintain only the first position for all players.

```{r}
# Removing lower case in nation
final_players$Nation <- gsub("[a-z]", "", final_players$Nation)
# Saving only the first position
final_players$Pos <- substr(final_players$Pos,1,2)
# Removing lower case before leaguew
final_players$Comp <- substr(final_players$Comp, 4, nchar(final_players$Comp))

# With that solved we will look at the players now to check for NAs and duplicated values
skim(final_players)
```

We see there are many NA values which could be from players who have barely played and therefore many of their data has not been registered. Because of this we will remove all players that have played less very little. For this we have to remove the columns for minutes played and number of starts of goalkeepers, and replace the X90s column in those.

```{r}
# Getting the expected 90s of goalkeepers in the same 
# column as the other players
#final_players$X90s <- final_players$X90s + final_players$X.90


# Eliminating players that have not played much
final_players <- final_players[final_players$X90s > 2,]

# Number of duplicated players 
sum(duplicated(final_players$Player))

skim(final_players)
```

Apart from that we also see there are a total of +200 duplicates which could be from players who were transferred or were loaned mid-season. To deal with this, we will keep the entries with the most time played.

```{r}
# Obtaining the data without repeated players
final_players <- final_players %>%
  group_by(Player) %>%
  # Keeping only observations with the most starts
  slice_max(Starts, n = 1, with_ties = FALSE) %>%  
  ungroup() %>%
  distinct()

# Check that there are no duplicates now
sum(duplicated(final_players))
```

Now we have to deal with NA values. Since all NA values are in numerical columns, we can replace these NAs by the median of that column.

```{r}
# Replace NA values with the median of each column
final_players <- final_players %>%
  mutate(across(everything(), ~ ifelse(is.na(.), median(., na.rm = TRUE), .)))

# Number of rows with NA values 
nrow(final_players %>% filter(if_any(everything(), is.na)))
```




We see that this dataset has many statistics for players, but we could also benefit from other ones that are not there, such as the overall player rating or their price. For this we will merge part of another dataset which bases its rating and some other statistics in the FIFA videogame.



```{r}
# Reading complimentary data and removing unnecessary columns
data_fifa <- read.csv("CLEAN_FIFA23_official_data.csv")
head(data_fifa)
skim(data_fifa)
colnames(data_fifa)
# Checking for duplicate values
sum(duplicated(data_fifa$Name))
``` 
As it has happened with our previous football datasets, we will have to modify duplicate values

```{r}
# Obtaining the data without repeted players
data_fifa <- data_fifa %>%
  group_by(Name) %>%
  # Keeping only observations with the most recent joined date
  slice_max(Joined, n = 1, with_ties = FALSE) %>%  
  ungroup() %>%
  distinct()

# Check that there are no duplicates now
sum(duplicated(final_players))




```

We can also see that the dataset is already clean with no Na values but looking at the different columns and variables  there are some columns which do not provide much information, are already in the previous dataset or are not of our interest for our current project, therefore, we will remove them.
Since the merging will be done by name it is the only column present in other datasets that we will maintain

```{r}
colnames(data_fifa)
cols <- c("X", "ID", "Age", "Photo","Flag", "Club", "Club.Logo", "Special","Work.Rate","Skill.Moves", "Weak.Foot","Body.Type", "Real.Face", "Position", "Joined", "Loaned.From", "Contract.Valid.Until",
          "Release.Clause...", "Kit.Number", "Best.Overall.Rating", "Year_Joined")

data_fifa <- data_fifa %>% select(-all_of(cols))



```


```{r}

# The format string for the players is different, therefore we will merge them according to the similarity in the player and the name
full_data <- stringdist_join(final_players, data_fifa,
                            by = c("Player" = "Name"),
                            method = "jw",
                            max_dist = 0.2,
                            distance_col = "dist")
names(full_data)

```

Now that we have cleaned a bit the other dataset we can proceed to merging it with this new and last one.

For this there is a problem, where FIFA players have different name formats ("Messi", "L. Messi" or "Lionel Messi"). To solve this we will first standardized name to the format "L. Messi" and then extract the surnames for both datasets and join by means of the surname.

Also we will replace before all of this all characters from other languages into the closest one in the English alphabet ("è" -\> "e")


#DEJE ESTE CHUNK POR SI ACASO, AHORA MISMO R NO LO LEE

```{r eval=FALSE, include=FALSE}
# Replacing characters by its closest English alphabet character
final_players$Player <- stri_trans_general(final_players$Player, "latin-ascii")
data_fifa$Name <- stri_trans_general(data_fifa$Name, "latin-ascii")

# Handle additional characters manually for specific languages 
final_players$Player <- gsub("ć|č", "c", final_players$Player)
final_players$Player <- gsub("ł", "l", final_players$Player)
final_players$Player <- gsub("ń", "n", final_players$Player)
final_players$Player <- gsub("ă", "a", final_players$Player)
final_players$Player <- gsub("ș|ş", "s", final_players$Player)
final_players$Player <- gsub("ž|ź", "z", final_players$Player)

# Removes other symbols
final_players$Player <- gsub("[^[:alnum:] ]", "", final_players$Player)  

data_fifa$Name <- gsub("ć|č", "c", data_fifa$Name)
data_fifa$Name <- gsub("ł", "l", data_fifa$Name)
data_fifa$Name <- gsub("ń", "n", data_fifa$Name)
data_fifa$Name <- gsub("ă", "a", data_fifa$Name)
data_fifa$Name <- gsub("ș|ş", "s", data_fifa$Name)
data_fifa$Name <- gsub("ž|ź", "z", data_fifa$Name)
data_fifa$Name <- gsub("[^[:alnum:] ]", "", data_fifa$Name)



# Set a Name column to the format "L. Messi"
data_fifa$StandardizedName <- paste0(substr(data_fifa$Name, 1, 1), ". ", sapply(strsplit(data_fifa$Name, " "), function(x) x[length(x)]))

# Create a new column of the surname
final_players$Surname <- sapply(strsplit(final_players$Player, " "), function(x) x[length(x)])
data_fifa$Surname <- sapply(strsplit(data_fifa$StandardizedName, " "), function(x) x[length(x)])

# Merge both datasets by Surname
full_data <- full_join(final_players, data_fifa, by = "Surname")

# Eliminating repeated values when merging has not been correct
full_data <- na.omit(full_data)
full_data <- distinct(full_data, Player, .keep_all = TRUE)
```

We observe that some observations has been lost. If we look closely at the data most of those observations not merged do not appear in the FIFA dataset so, since they are not very relevant we can exclude them confidently.

```{r}
# Observations not merged
not_merged <- anti_join(final_players, full_data, by = "Player")
```

Now that we have the whole data, we will address the huge amount of columns we have. First of all, we will remove columns that reference the same statistics, and also get rid of the useless ones.

```{r}
# Getting rid of useless and repeated columns
useless_repeated_cols = c("Year_Joined", "Best.Overall.Rating", "Kit.Number", "Release.Clause...", "Contract.Valid.Until", "Loaned.From", "Joined", "Position", "Real.Face", "Body.Type", "Work.Rate", "Skill.Moves", "Weak.Foot", "International.Reputation", "Special", "Wage...", "Club.Logo", "Club", "Potential", "Flag", "Nationality", "Photo", "Age.y", "Name", "ID" )

full_data <- full_data %>% select(-all_of(useless_repeated_cols))


```
