---
title: "R Notebook"
output: html_notebook
---

**SET UP**

-   Importing libraries

-   Importing initial data

```{r}
rm(list = ls())

# Importing libraries
library("dplyr")
library("skimr")

# Reading data
#data <- read.csv("2022-2023 Football Player Stats.csv", sep=";")
data <- read.csv("top5-players.csv")
```

**DATA PREPROCESSING**

Getting an idea of what the dataset looks like.

```{r}
skim(data)
```

As we observe from the output there are a couple of NA values, so we have to decide what to do with them.

```{r}
# Select rows with any NA values
data[!complete.cases(data), ]
```

The first player has many stats missing and since he has only played 1 game in the season we will delete it. The next ones have their age, and born (year) missing (one also has its country missing), but with a quick Google search we can manually input these variables.

```{r}
# Eliminating player with many missing variables
data <- data %>% filter(Rk != 1639)

# Mnaually imputing some missing values
data[data$Rk == 1785, c("Age", "Born")] <- c("17", "2006")
data[data$Rk == 2075, c("Nation", "Age", "Born")] <- c("ar ARG", "21", "2002")
data[data$Rk == 2076, c("Nation", "Age", "Born")] <- c("ar ARG", "21", "2002")
data[data$Rk == 2530, c("Age", "Born")] <- c("21", "2001")

# Select rows with any NA values
data[!complete.cases(data), ]
```

Now there are no rows with missing values, but from the players with NA values we have discovered that there are some players that appear more than once. This might be because maybe they played a couple of games for one team and then got transferred or loaned to another one.

In this case what we will do is combine all the instances of that player in the dataset into a single observation and recalculate all statistics based on the combination of the different rows.

```{r}
# Obtaining the data without repetition
unique_data <- data %>%
  group_by(Player) %>%
  summarise(
    # Categorical variables
    Nation = first(Nation), 
    Pos = first(Pos),
    Age = first(Age), 
    Born = first(Born),
    # Selecting team and competition where
    # the player has the most starts
    Squad = Squad[which.max(Starts)],  
    Comp = Comp[which.max(Starts)],
    
    
    # Sum the numeric columns since all have 
    # linear statistics of the players
    across(where(is.numeric), sum),
    
    .groups = "drop"
  )

skim(unique_data)
```
